---
title: "AlcoChallenge"
author: "Helene Bæk Juel"
date: "6/29/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load packages
source(here::here("R/package-loading.R"))
```

## Load full dataset
```{r load-data}
# Load data from raw NPX data
olink <- OlinkAnalyze::read_NPX(here::here("data-raw/20202249_Juel_NPX.xlsx"))

# Fix issue with TIPS naming
# replace typo TIEP80EC	--> TIPS80EC
olink2 <- olink %>%
    mutate(SampleID = sub("TIEP80EC", "TIPS80EC", SampleID))
# replace typo TIPSKC/93 --> TIPS93
olink2 <- olink2 %>%
    mutate(SampleID = sub("TIPSKC/93", "TIPS93", SampleID))

# Add columns for cohort and pt_ID to be able to extract each cohort
olink2 <- olink2 %>%
    mutate(cohort = sub("[0-9]{1,4}.*$", "", SampleID)) %>% # substitute 1-4 numbers and everything after this (.*$) with nothing
    mutate(cohort = sub("_.*$", "", cohort)) # substitute anything after a _ with nothing to remove long control names and just keep "CONTROL"

# change cytokine names
# first remove all -
# then change spaces to _
# then rename alpha/beta/gamma to a/b/g
olink3 <- olink2 %>%
    mutate(Assay = gsub("-", "", Assay)) %>%
    mutate(Assay = gsub(" ", "_", Assay)) %>%
    mutate(Assay = gsub("alpha", "a", Assay)) %>%
    mutate(Assay = gsub("beta", "b", Assay)) %>%
    mutate(Assay = gsub("gamma", "g", Assay))

# change values < LOD of that assay to 50% LOD
olink4 <- olink3 %>%
    mutate(corr_NPX = if_else(LOD > NPX, LOD/2, NPX))
```

# Wrangling and QC

## Subset Alcochallenge datapoints
```{r QC1}
# ALCO
# Create columns for pt_ID, time_point, sample_type for each cohort
ALCO <- olink4 %>%
    filter(cohort == "ALCO") %>%
    mutate(pt_ID = sub("^[A-Z]{2,7}", "", sub("-.+$", "", SampleID))) %>%  # substitute 2-7 letters and anything after a "-" with nothing
    mutate(time_point = sub("^.+-T", "", sub("S.+$", "", SampleID))) %>% # add timepoint
    mutate(sample_type = sub("^.+S", "", SampleID)) # add sample_type (peripheral or liver vein)

# Remove duplicate sample
# ALCO %>%
#     filter(sample_type == "1-d") #92 duplicates (1 sample)
ALCO <- ALCO %>%
    filter(sample_type != "1-d") # goes from 22172 to 22080 obs

```


## Visualize potential outliers by IQR vs. sample median
```{r QC_plot}
olink_qc_plot(ALCO)
```

6 samples have a QC warning and 2 appear to be outliers

## Further QC
```{r QC2}
# remove samples with QC warning
# ALCO_fail <- ALCO %>%
#     filter(QC_Warning != "Pass") # 552 obs = 6 samples
# ALCO_fail %>%
#     filter(pt_ID != "106") - all QC warnings are from pt 106
ALCO <- ALCO %>%
    filter(QC_Warning == "Pass") # goes from 22080 to 21528 obs


# Add subgroup column: 100=ALD, 200=NAFLD, 300=healthy
ALCO$pt_ID <- as.numeric(ALCO$pt_ID)
ALCO <- ALCO %>%
    mutate(subgroup = if_else(pt_ID > 300, "Healthy", if_else(pt_ID > 200, "NAFLD", "ALD")))

# Change pt_ID, time_point, sample_type and subgroup to factor
ALCO$pt_ID <- as.factor(ALCO$pt_ID) # 39 indv - ID106 removed in QC warning
ALCO$time_point <- as.factor(ALCO$time_point) # 3 TPs
ALCO$sample_type <- as.factor(ALCO$sample_type) # 2 sample types
ALCO$subgroup <- as.factor(ALCO$subgroup) # 3 subgroups

# Change order of subgroups
ALCO$subgroup <- ordered(ALCO$subgroup, levels = c("Healthy", "NAFLD", "ALD"))

```

# Results

## Plot all cytokines by patient subgroup (not included)
```{r plot_all, include = F}
# Olink ID list
# cytokines <- levels(as.factor(olink4$Assay)) # protein abbreviations
cytokine_IDs <- levels(as.factor(olink4$OlinkID)) #Olink IDs

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = cytokine_IDs[1:16],
              verbose = T,
              number_of_proteins_per_plot = 16)

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = cytokine_IDs[17:32],
              verbose = T,
              number_of_proteins_per_plot = 16)

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = cytokine_IDs[33:48],
              verbose = T,
              number_of_proteins_per_plot = 16)

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = cytokine_IDs[49:64],
              verbose = T,
              number_of_proteins_per_plot = 16)

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = cytokine_IDs[65:80],
              verbose = T,
              number_of_proteins_per_plot = 16)

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = cytokine_IDs[81:92],
              verbose = T,
              number_of_proteins_per_plot = 12)

```

## ANOVA by patient subgroup
```{r ANOVA_subgroup}
anova_ALCO_group <- olink_anova(ALCO, variable = "subgroup")

sign_cytokines <- anova_ALCO_group %>%
    filter(Adjusted_pval < 0.0001) %>%
    pull(Assay)

print(sign_cytokines)
```

## Plot significant cytokines by patient subgroup
```{r plot_subgroup}
# Extract significant protein IDs
sign_IDs <- anova_ALCO_group %>%
    filter(Adjusted_pval < 0.0001) %>%
    pull(OlinkID)

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = sign_IDs[1:9],
              verbose = T,
              number_of_proteins_per_plot = 9)

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = sign_IDs[10:18],
              verbose = T,
              number_of_proteins_per_plot = 9)

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = sign_IDs[19:27],
              verbose = T,
              number_of_proteins_per_plot = 9)

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = sign_IDs[28:36],
              verbose = T,
              number_of_proteins_per_plot = 9)

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = sign_IDs[37:45],
              verbose = T,
              number_of_proteins_per_plot = 9)

olink_boxplot(ALCO,
              variable = "subgroup",
              olinkid_list = sign_IDs[46:52],
              verbose = T,
              number_of_proteins_per_plot = 7)
```



## ANOVA by sample type
```{r ANOVA_sample}
anova_ALCO_sample <- olink_anova(ALCO, variable = "sample_type")

```

## Plot significant cytokines by sample type
```{r plot_sample}
# Extract significant protein IDs
sign_IDs <- anova_ALCO_sample %>%
    filter(Adjusted_pval < 0.001) %>%
    pull(OlinkID)

olink_boxplot(ALCO,
              variable = "sample_type",
              olinkid_list = sign_IDs,
              verbose = T,
              number_of_proteins_per_plot = 5)
```


## ANOVA by timepoint in liver vein
TODO: check sample_type naming
```{r ANOVA_liver_TP}
ALCO_livervein <- ALCO %>% 
    filter(sample_type == 2)

anova_ALCO_liver_TP <- olink_anova(ALCO_livervein, variable = "time_point")

```
## Plot significant cytokines by timepoint in liver vein
```{r plot_liver_TP}
# Extract significant protein IDs
sign_IDs <- anova_ALCO_liver_TP %>%
    filter(Adjusted_pval < 0.001) %>%
    pull(OlinkID)

olink_boxplot(ALCO_livervein,
              variable = "time_point",
              olinkid_list = sign_IDs,
              verbose = T,
              number_of_proteins_per_plot = 3)
```

## ANOVA by timepoint in peripheral blood
```{r ANOVA_peripheral_TP}
ALCO_peripheral <- ALCO %>% 
    filter(sample_type == 1)

anova_ALCO_peripheral_TP <- olink_anova(ALCO_peripheral, variable = "time_point")

```
## Plot significant cytokines by timepoint in peripheral blood
```{r plot_peripheral_TP}
# Extract significant protein IDs
sign_IDs <- anova_ALCO_peripheral_TP %>%
    filter(Adjusted_pval < 0.001) %>%
    pull(OlinkID)

olink_boxplot(ALCO_peripheral,
              variable = "time_point",
              olinkid_list = sign_IDs,
              verbose = T,
              number_of_proteins_per_plot = 2)
```

### The following packages and versions were used to create this report
```{r sessioninfo}
sessionInfo()
```


### This document was knitted on
```{r knitdate}
Sys.Date()
```




