---
title: "GALA_olink"
author: "Helene Bæk Juel"
date: "9/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load packages
source(here::here("R/package-loading.R"))
```

# Load, wrangle and QC the three datasets: Olink, phenotypes and genetics
The entire section can be skipped if loading the combined dataset directly (GALA.rda).

## Load full Olink dataset
Can be skipped (include=FALSE) if olink3/4 loaded directly.
```{r load-olink}
# Load data from raw NPX data
olink <- OlinkAnalyze::read_NPX(here::here("data-raw/20202249_Juel_NPX.xlsx"))

# Fix issue with TIPS naming
# replace typo TIEP80EC	--> TIPS80EC
olink2 <- olink %>%
    mutate(SampleID = sub("TIEP80EC", "TIPS80EC", SampleID))
# replace typo TIPSKC/93 --> TIPS93
olink2 <- olink2 %>%
    mutate(SampleID = sub("TIPSKC/93", "TIPS93", SampleID))

# Add columns for cohort and pt_ID to be able to extract each cohort
olink2 <- olink2 %>%
    mutate(cohort = sub("[0-9]{1,4}.*$", "", SampleID)) %>% # substitute 1-4 numbers and everything after this (.*$) with nothing
    mutate(cohort = sub("_.*$", "", cohort)) # substitute anything after a _ with nothing to remove long control names and just keep "CONTROL"

# change cytokine names
# first remove all -
# then change spaces to _
# then rename alpha/beta/gamma to a/b/g
olink3 <- olink2 %>%
    mutate(Assay = gsub("-", "", Assay)) %>%
    mutate(Assay = gsub(" ", "_", Assay)) %>%
    mutate(Assay = gsub("alpha", "a", Assay)) %>%
    mutate(Assay = gsub("beta", "b", Assay)) %>%
    mutate(Assay = gsub("gamma", "g", Assay))

# change values < LOD of that assay to 50% LOD
olink4 <- olink3 %>%
    mutate(corr_NPX = if_else(LOD > NPX, LOD/2, NPX))
# TODO: consider whether this is needed - and if so, overwrite NPX (used in Olinkanalyze package)
```

## Filter GALA samples
```{r QC1-olink}
# Load dataset if not running "Load full dataset" chunk above
# load(here::here("data/olink4.rda"))

GALA_olink <- olink3 %>% # olink4 has values below LoD substituted with 50% LoD. Olink3 uses the measured NPX values
    filter(cohort == "ALD" | cohort == "HP") %>%
    mutate(sample_type = sub("^.+[0-9]{1,3}", "", SampleID)) # duplicate sample marked as "-d"

# Remove duplicate samples
GALA_olink <- GALA_olink %>%
    filter(sample_type != "-d")  # goes from 50692 to 50416 obs

GALA_olink$sample_type <- NULL # remove sample_type column

# TODO: Investigate why we only have 548 GALA samples (expected ~463+138 = 600)
```


## Visualize potential outliers by IQR vs. sample median
```{r QC_plot}
olink_qc_plot(GALA_olink)
```

Some samples have a QC warning and 2 appear to be outliers

## Further QC
```{r QC2-olink}
# remove samples with QC warning
GALA_olink <- GALA_olink %>%
    filter(QC_Warning == "Pass") # goes from 50416 to 49404 obs

# Remove extra 0's in SampleID HP004-HP096 to be able to merge with phenotype data
GALA_olink <- GALA_olink %>%
  mutate(SampleID = sub("^HP[0]{1,2}", "HP", SampleID))

# Remove cohort column before merging with phenotype data
GALA_olink$cohort <- NULL

```

## Load and wrangle phenotype data
```{r load-phenotype, warning=FALSE}
# TODO: Change to new QC'ed phenotype dataset
GALA_pheno <- read_excel(here::here("data-raw/ALD_HP_outcome_HBJ.xlsx"))
str(GALA_pheno)

# Change data types
GALA_pheno2 <- GALA_pheno %>% 
  mutate(kleiner_numeric = kleiner, 
         steatosis_numeric = nas_steatosis, 
         inflam_numeric = nas_inflam) %>% #Add numeric columns (before parent column is converted to factor) 
  mutate(across(c(cpa, meld, elf, te, # liver scores
                  packyears, 
                  height, weight, bmi, waist, hip, whr, # anthropometry
                  hr, map, sbp, dbp, # heart
                  alt, ast, alk, bili, crp, ggt, # liver enzymes
                  chol, hdl, ldl, trigly, # lipids
                  iga, igg, igm, leu,  # antibodies
                  glc, hba1c, insulin, homair, cpeptid, # glucose metabolism
                  proc3, 
                  days_to_LRE, days_to_hospInf, days_to_mort, 
                  kleiner_numeric, steatosis_numeric, inflam_numeric), as.numeric)) %>% # Change data columns to numeric
    mutate(te_fibrosis_6 = if_else(te <= 6, "low", "high"),
           te_fibrosis_8 = if_else(te <= 8, "low", "high"), # Add columns for fibrosis level high or low based on TE
           steatosis_binary = if_else(steatosis_numeric < 2, "low", "high")) %>% 
# TODO: also add binary inflammation and steatosis
  mutate(across(c(cohort, gender, 
                  kleiner, fibrosis, starts_with("nas"), inflam, # liver scores
                  abstinent, overuse, alcoholyears, abstinenceyears, # alcohol
                  liverrelated_event, hospInfection, allMortality, # outcomes
                  te_fibrosis_6, te_fibrosis_8, steatosis_binary,
                  excess_drink_followup), factor)) %>% # Change data columns to factor
# TODO: order factors alcoholyears and abstinenceyears
  mutate(kleiner_numeric = if_else(is.na(kleiner_numeric), 0.5, kleiner_numeric)) # assign NA to 0.5, since those without a kleiner score are known to have no or low fibrosis
```
## Compute new endpoints from follow-up data
```{r QC-phenotype, warning=FALSE}
# Add more endpoint columns 
GALA_pheno3 <- GALA_pheno2 %>%  
  mutate(days_to_hospInf2 = dplyr::if_else(is.na(days_to_hospInf), ifelse(allMortality == "no", 2*max(days_to_hospInf, na.rm = T), NA), days_to_hospInf), # If the patient did not have a hospInf (n=321), assign 2x max(days_to_hospInf) to pts who did not die and NA to those who died without having an infection (n=31). Those that were not followed up (n=155) retain NA. 
         #Note that dplyr::if_else() requires T and F values to be of the same type, hence you cannot assign NA together with a numeric vector. baseR ifelse() does not have this constraint.
         days_to_LRE2 = dplyr::if_else(is.na(days_to_LRE), ifelse(allMortality == "no", 2*max(days_to_LRE, na.rm = T), NA), days_to_LRE), # no LRE n=378, no follow-up n=155
         days_to_mort2 = dplyr::if_else(allMortality == "no", 2*max(days_to_mort, na.rm = T), ifelse(allMortality == "yes", days_to_mort, NA)), # no death n=386, no follow-up n=155
         hospInf_1yr = as.factor(ifelse(days_to_hospInf2 == "NA", NA, if_else(days_to_hospInf2 < 366, "yes", "no"))), # NA = not followed up or died without having an infection (n=186); yes = the pt was followed up AND had an infection in the first year (n=39); no = the pt was followed up AND DID NOT have an infection in the first year (n=392)
         LRE_1yr = as.factor(ifelse(days_to_LRE2 == "NA", NA, if_else(days_to_LRE2 < 366, "yes", "no"))), # n=33
         mort_1yr = as.factor(ifelse(days_to_mort2 == "NA", NA, if_else(days_to_mort2 < 366, "yes", "no"))), # n=11
         any_event_tot = as.factor(if_else(c(hospInfection == "yes" | liverrelated_event == "yes" | allMortality == "yes"), "yes", ifelse(liverrelated_event == "NA", NA, "no"))), # n=184
         any_event_1yr = as.factor(if_else(c(hospInf_1yr == "yes" | LRE_1yr == "yes" | mort_1yr == "yes"), "yes", ifelse(c(liverrelated_event == "NA" | hospInfection == "NA" | allMortality == "NA"), NA, "no")))) # n=65
# summary(GALA_pheno3[, 54:74])

# Change column name from CBMR_ID to SampleID to allow merging with olink data
GALA_pheno3 <- GALA_pheno3 %>%
  rename(SampleID = CBMR_ID)

# Change order of factors, if needed
GALA_pheno3$cohort <- ordered(GALA_pheno3$cohort, levels = c("HP", "ALD"))

# save dataset in data folder
# usethis::use_data(GALA_pheno3, overwrite = T)
```

## Match and merge Olink with phenotype data
```{r merge1}
# Match and merge Olink with phenotype data
GALA0 <- merge(GALA_olink, GALA_pheno3, by = "SampleID") # goes from 49404 to 49312 obs from GALA_olink = 1 ppt!

# Find missing sample in phenotype data
print(setdiff(unique(GALA_olink$SampleID), GALA_pheno3$SampleID)) # ALD1302

# Find missing samples in Olink data
print(setdiff(unique(GALA_pheno3$SampleID), GALA_olink$SampleID)) # 81 missing samples

```
TODO: replace phenotype dataset with final QC'd version
TODO: add phenotype data for ALD1302
TODO: investigate 81 missing Olink samples (59 ALD and 22 HP)

## Load genetic data to add genotype SNP and PRS dataset
```{r load-genotype}
# Add PRS and SNP data
ALD_gen <- read_excel(here::here("data-raw/genALD_HBJ.xlsx"), sheet = "combined") #489 obs
HP_gen <- read_excel(here::here("data-raw/genHP_HBJ.xlsx"), sheet = "combined") #136 obs
GALA_gen <- rbind(ALD_gen, HP_gen) #625 obs
# Change column name from CBMR_ID to SampleID to allow merging with olink data
GALA_gen <- GALA_gen %>%
  rename(SampleID = CBMR_ID)

# save dataset in data folder
# usethis::use_data(GALA_gen, overwrite = T)
```

## Merge genotypes with Olink+phenotype data
```{r merge2}
GALA <- merge(GALA0, GALA_gen, by = "SampleID") # goes from 49312 to 48852 obs from GALA0 = 5 ppts
print(setdiff(unique(GALA0$SampleID), GALA_gen$SampleID)) # ALD2265, ALD2332, ALD2547, ALD2558, ALD402

# save dataset in data folder
# usethis::use_data(GALA, overwrite = T)
```
TODO: investigate why we are missing genotypes for these 5.

## Load merged dataset
If not loading and merging individual datasets
```{r load_merged}
# Load dataset if not running the loading and merging chunks above
# load(here::here("data/GALA.rda"))
```


# Results

## Plot all cytokines vs Kleiner score
Not included in current knit (include=FALSE).
```{r plot_all_kleiner, warning=FALSE}
# Olink ID lists for looping
cytokines <- levels(as.factor(olink3$Assay)) # protein abbreviations
cytokine_IDs <- levels(as.factor(olink3$OlinkID)) #Olink IDs

# Loop over all cytokines
allcytokines_kleiner <- list()

for(i in cytokines[1:92]) {
    ploti <- GALA %>%
        # filter(!is.na(as.numeric(kleiner))) %>% #to plot only those with biopsy
        filter(Assay == i) %>%
        ggplot(aes(x = kleiner, y = NPX, color = cohort)) +
        geom_violin(draw_quantiles = c(0.5)) +
        geom_jitter(position = position_jitterdodge(), size = 0.5) + # jitterdodge needs groups to dodge (color = cohort)
        ggtitle(paste(i))

    allcytokines_kleiner[[i]] <- ploti
    print(ploti)
}

# To save plotgrid as pdf
# pdf(here::here("doc/images/GALA_allcytokines_kleiner.pdf"), height = 60, width = 40) # height = 5 for each row of 8 -> 60 for 12x8
# do.call('grid.arrange', c(allcytokines_kleiner, ncol=8))
# dev.off()
```

## Run ANOVA for Kleiner score

Plots are ordered by ascending p-val.
```{r ANOVA_kleiner}
# count n
# summary(as.factor(GALA$kleiner)) # F0: 32, F1: 114, F2: 88, F3: 20, F4: 56, NA: 226. Sum !=NA: 310

# run ANOVA
anova_GALA_kleiner <- olink_anova(GALA, variable = "kleiner")

# Extract list of significant cytokines
sign_cytokines <- anova_GALA_kleiner %>%
    filter(Adjusted_pval < 0.05) %>% 
    pull(Assay)
print(sign_cytokines)
```
84 of the 92 cytokines have p<0.05 after Benjamini-Hochberg correction for multiple testing.

## Plot significant cytokines (Kleiner)
Not included in current knit if include=FALSE.
```{r plot_sign_kleiner, warning=FALSE, include=FALSE}
# Prepare list for graphs
signcytokines_kleiner <- list()

for(i in sign_cytokines[1:72]) {
    ploti <- GALA %>%
        # filter(!is.na(as.numeric(kleiner))) %>% #to plot only those with biopsy
        filter(Assay == i) %>%
        ggplot(aes(x = kleiner, y = NPX, color = cohort)) +
        geom_violin(draw_quantiles = c(0.5)) +
        geom_jitter(position = position_jitterdodge(), size = 0.5) + # jitterdodge needs groups to dodge (color = cohort)
        ggtitle(paste(i))

    signcytokines_kleiner[[i]] <- ploti
    print(ploti)
}

# To save plotgrid as pdf
# pdf(here::here("doc/images/GALA_signcytokines_kleiner.pdf"), height = 45, width = 40) # height = 5 for each row of 8 -> 45 for 9x8
# do.call('grid.arrange', c(signcytokines_kleiner, ncol=8))
# dev.off()
```
Observations from plots:

- Most cytokines (except SCF) increase with increasing Kleiner score
- Some cytokines appear to increase already from Kleiner 1 to 2 (HGF, IL8, CCL19, CDCP1, LAP-TGFb, others)
- Some cytokines appear to increase only with high fibrosis (SLAMF1, uPA, IL15RA, PDL1, CCL28, CD8A TNFb, others)
- Need to consider how to analyse cytokines with many observations below LoD (e.g. IL6, FGF23, GDNF, SIRT2, ST1A1, MCP3, IL17A, ARTN)
- FGF21 appears to have a bell-shaped curve, with maximum at Kleiner 2. Interestingly, a significant increase in FGF21 was observed 24h(? TP3) after alcohol bolus in AlcoChallenge, in both liver vein and peripheral blood. No other cytokines showed significant increase (HGF and NRTN decreased).

Other interesting findings to consider:
- IL6, OSM, CXCL10, CCL19, MCP4 all significantly higher in liver vein compared with peripheral blood in AlcoChallenge study --> being produced by the liver? 

## Linear model for TE
```{r lm_te}
# count n
# summary(GALA$te) # 18 NA

lm_GALA_te <- olink_lmer(GALA, 
           variable = "te", 
           outcome = "NPX",
           random = "cohort")

# Extract list of significant cytokines
sign_cytokines <- lm_GALA_te %>%
    filter(Adjusted_pval < 0.05) %>% 
    pull(Assay)
print(sign_cytokines) # 68 cytokines significant 
```
68 cytokines have p<0.05 after Benjamini-Hochberg correction for multiple testing.

## Plot cytokines with significant linear correlation with TE
```{r plot_sign_TE, warning=FALSE}
# Prepare list for graphs
signcytokines_te <- list()

for(i in sign_cytokines[1:68]) {
    ploti <- GALA %>%
        filter(Assay == i) %>%
        ggplot(aes(x = te, y = NPX)) +
        geom_point(aes(color = cohort)) +
        geom_smooth(method = lm, color = "black") +
        scale_x_log10() +
        ggtitle(paste(i))

    signcytokines_te[[i]] <- ploti
    print(ploti)
}

# To save plotgrid as pdf
# pdf(here::here("doc/images/GALA_signcytokines_te.pdf"), height = 45, width = 40) # height = 5 for each row of 8 -> 45 for 9x8
# do.call('grid.arrange', c(signcytokines_te, ncol=8))
# dev.off()
```

- In addition to SCF, TRAIL, CXCL5, CCL23 also show negative correlation with TE


## Linear model for outcome
TODO: Fix Error: grouping factors must have > 1 sampled level
Maybe not needed after Lili runs their analysis.
```{r lm_death, eval=FALSE, include=FALSE}
# count n
# summary(GALA$days_to_LRE2) # 18 NA

lm_GALA_LRE <- olink_lmer(GALA,
           variable = "days_to_LRE2",
           outcome = "NPX",
           random = "cohort")

# Extract list of significant cytokines
sign_cytokines <- lm_GALA_LRE %>%
    filter(Adjusted_pval < 0.05) %>% # adjusted pval already adjusted with Benjamini-Hochberg
    pull(Assay)
print(sign_cytokines) # xx cytokines significant
```


### The following packages and versions were used to create this report
```{r sessioninfo}
sessionInfo()
```


### This document was knitted on
```{r knitdate}
Sys.Date()
```

